### 2023-02-11 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_interpreting_probabilities@6f55f70290a06f30abd441eb8a2cae1bc8399fe1 🚀

The basic unit of food webs and other ecological networks are
individuals that interact with each others (e.g., by predation; Elton
2001), forming individual-based networks. The aggregation of these
individuals into more or less homogeneous groups (e.g., populations,
species, trophic species, families) allows us to represent networks at
broader scales, which impacts the properties and behaviour of these
systems (Guimarães 2020). A network's nodes can thus designate distinct
levels of organization, whereas the edges linking these nodes can
describe a variety of interaction measures. When using a Boolean
(yes-no) representation of biotic interactions, the observation that one
individual from group (or node) \(i\) interacts with another individual
from group \(j\) is enough to set the interaction \(A_{i,j}\) to 1. This
simplified representation of food webs is a highly valuable source of
ecological information (Pascual \emph{et al.} 2006) even though it
overlooks important factors regarding interaction strengths. These, in
turn, can be represented using weighted interactions, which better
describe the energy flows, demographic impacts or frequencies of
interactions between nodes (Berlow \emph{et al.} 2004; Borrett \&
Scharler 2019), with \(A_{i,j} \in \mathbb{N}\) or \(\mathbb{R}\)
depending on the measure. For example, they can be used to estimate the
average number of prey individuals consumed by the predators in a given
time period. This extra amount of ecological information typically comes
at a cost of greater sampling effort or data requirement in predictive
models (Strydom \emph{et al.} 2021), which can lead to high
uncertainties when building these types of networks. Therefore,
important methodological and conceptual decisions must been made when
sampling and building food webs.

@@

Starting from a selected set of species, metawebs can be built using
different data sources, including literature review, field work, and
predictive models (e.g., the metaweb of Canadian mammals inferred by
Strydom \emph{et al.} 2022). Every pair of species that has been
observed to interact at least once can be given a probability of
interaction of \(1\); we know that they \emph{can} interact. This means
that rare interactions can technically be given high probabilities in
the metaweb. Unobserved interactions, on the other hand, are given lower
probabilities, going as low as \(0\) for forbidden links. Two important
nuances must however be made here. Because of observation errors due to
taxonomic misidentifications and ecological misinterpretations, many
observations of interactions are actually false positives. Similarly,
forbidden interactions can be false negatives if e.g.~they have been
assessed for specific phenotypes, locations or time. Implementing a
Bayesian framework, which updates prior probabilities of interactions
with empirical data, could lessen these errors.

@@

As opposed to metawebs, probabilistic local food webs represent the
likelihood that two species will interact at a specific location and
within a given time period; in other words, they are context dependant.
To be specific, space is defined here as the geographic coordinates
\(x\), \(y\) of the spatial boundaries delineating the system, whereas
time is the time interval \(t\) during which interactions were sampled
or for which they were predicted. We want to point out that they are not
single values, but rather continued dimensions that could be outlined
differently depending on the study system. Regardless of how they were
defined, they always delineate a specific area \(A\) and duration \(D\).
These could refer to the sampled area and duration or to the targeted
location and time period.

Many factors could be taken into consideration when estimating the
probability that a predator species \(i\) interacts with a given prey
species \(j\) locally. One of the most important is species
co-occurrence, which describes if both species can be found at location
and time \(x\), \(y\), \(t\). Surely, the probability that the
interaction is realized must be \(0\) when species do not co-occur.
Similarly, species relative abundance is another important predictor of
the probability of interaction, because it impacts the probability that
species will randomly encounter (Canard \emph{et al.} 2012; Canard
\emph{et al.} 2014; Poisot \emph{et al.} 2015). Interaction
probabilities can also be conditional on other biological and
environmental variables, such as temperature, precipitation, presence of
shelters, phenotypic plasticity, phenology, and presence of other
interacting species in the network. These conditions can affect species
traits, which greatly impact the likelihood of interactions (Poisot
\emph{et al.} 2015). When probabilities of interactions are estimated
using specific values of these variables, ecologists should make them
explicit in their metadata, preferably using mathematical equations to
avoid any ambiguity. In the next section, we will see examples of why
this matters when it comes to spatial, temporal, and taxonomic scaling
of biotic interactions. Here, we will use the variable \(\Omega\) as a
substitute for the biological and ecological context in which
interaction probabilities were estimated. This gives us the following
equation for the probability that species (or taxa) \(i\) and \(j\)
interact in a local network \(N\):

\begin{equation}\protect\hypertarget{eq:local}{}{P_N(i \rightarrow j | x, y, t, \Omega).}\label{eq:local}\end{equation}

@@

Even though the spatiotemporal variation of interactions is not
considered in metawebs, they can still be useful to reconstruct local
networks of realized interactions. Indeed, local networks are formed
from subsets of their metaweb (called subnetworks), which are obtained
by selecting a subset of both species and interactions (Dunne 2006).
Because a community's composition is arguably easier to sample (or
predict) than its interactions, the biggest challenge is to sample links
from the metaweb. This becomes a conceptual issue when we consider how
potential and realized interactions differ. Despite these concerns,
metawebs remain an important source of ecological information that can
be leveraged for realistically predicting spatially explicit networks.
First, metawebs set the upper limit of species interactions (McLeod
\emph{et al.} 2021), i.e. the probability that two species interact at a
specific location is always lower or equal to the probability of their
potential interaction:

\begin{equation}\protect\hypertarget{eq:switch}{}{P_N(i \rightarrow j | x, y, t, \Omega) \le
P_M(i \rightarrow j).}\label{eq:switch}\end{equation}

Therefore, building local networks from their metaweb keeping the same
values of interaction probability would generate systematic biases in
the prediction. Second, the structure of local networks is constrained
by the one of their metaweb (Saravia \emph{et al.} 2022). This suggests
that a metaweb not only constraints the pairwise interactions of its
corresponding local networks, but also their emerging properties.
Inferring the structure of local networks from the metaweb could thus
help estimate more realistically the likelihood that potential
interactions are realized and observed locally (Strydom \emph{et al.}
2021).

@@

\begin{itemize}
\tightlist
\item
  Why do probabilistic local food webs scale with area and time but not
  probabilistic metawebs?
\item
  In metawebs, interaction probabilities do not scale with space and
  time because they depend solely on the biological capacity of two
  species to interact.
\item
  In local food webs, interaction probabilities scale with space and
  time because there are more opportunities of interactions (e.g., more
  environmental conditions) in a larger area and longer time period.
\item
  What are some network area relationships in probabilistic local food
  webs?
\item
  We know that local networks can inform regional networks. However, can
  regional networks inform local networks?
\end{itemize}

@@

\hypertarget{concluding-remarks}{%
\section{Concluding remarks}\label{concluding-remarks}}

Here we present some advice moving forward.

\begin{itemize}
\tightlist
\item
  What can we learn from other systems/fields (e.g., social networks,
  probabilistic graph theory)?
\item
  What even are probabilities? What is the probability that we will ever
  know the answer to that?
\item
  Be careful of how we define probabilities. Be sure to be explicit
  about these things. Be sure to specify the type of interaction, the
  spatial, temporal, and taxonomic scale when presenting new data on
  interaction probabilities. We need better metadata documentation.
\item
  Be careful to use and manipulate interaction probabilities properly
  depending on how they were defined and obtained. Different
  interpretations imply different scaling, and thus different ways to
  manipulate these numbers.
\item
  Maybe mention thinking about a workflow to predict probabilistic local
  food webs from probabilistic metawebs.
\end{itemize}

@@

>>>>

The basic unit of food webs and other ecological networks are
individuals that interact with each others (e.g., by predation; Elton
2001), forming individual-based networks. The aggregation of these
individuals into more or less homogeneous groups (e.g., populations,
species, trophic species, families) allows us to represent networks at
broader scales, which impacts the properties and behaviour of these
systems (Guimarães 2020). A network's nodes can thus designate distinct
levels of organization, whereas the edges linking these nodes can
describe a variety of interaction measures. When using a Boolean
(yes-no) representation of biotic interactions, the observation that one
individual from group (or node) \(i\) interacts with another individual
from group \(j\) is enough to set the interaction \(A_{i,j}\) to 1. This
simplified representation of food webs is a highly valuable source of
ecological information (Pascual \emph{et al.} 2006) even though it
overlooks important factors regarding interaction strengths. These, in
turn, can be represented using weighted interactions, which better
describe the energy flows, demographic impacts or frequencies of
interactions between nodes (Berlow \emph{et al.} 2004; Borrett \&
Scharler 2019), with \(A_{i,j} \in \mathbb{N}\) or \(\mathbb{R}\)
depending on the measure. For example, they can be used to estimate the
average number of prey individuals consumed by the predators in a given
time period. This extra amount of ecological information typically comes
at a cost of greater sampling effort or data requirement in predictive
models (Strydom \emph{et al.} 2021), which can lead to high
uncertainties when building these types of networks. Therefore,
important methodological and conceptual decisions must be made when
sampling and building food webs.

@@

Starting from a selected set of species, metawebs can be built using
different data sources, including literature review, field work, and
predictive models (e.g., the metaweb of Canadian mammals inferred by
Strydom \emph{et al.} 2022). Every pair of species that has been
observed to interact at least once can be given a probability of
interaction of \(1\); we know that they \emph{can} interact. This means
that rare interactions can technically be given high probabilities in
the metaweb. Unobserved interactions, on the other hand, are given lower
probabilities, going as low as \(0\) for forbidden links. Two important
nuances must however be made here. Because of observation errors due to
taxonomic misidentifications and ecological misinterpretations (e.g.,
due to cryptic species and interactions; Pringle \& Hutchinson 2020),
many observations of interactions are actually false positives.
Similarly, forbidden interactions can be false negatives if e.g.~they
have been assessed for specific phenotypes, locations or time.
Implementing a Bayesian framework, which updates prior probabilities of
interactions with empirical data, could lessen these errors.

@@

As opposed to metawebs, probabilistic local food webs represent the
likelihood that two species will interact at a specific location and
within a given time period; in other words, they are context dependant.
To be specific, space is defined here as the geographic coordinates
\((x, y)\) of the spatial boundaries delineating the system, whereas
time is the time interval \(t\) during which interactions were sampled
or for which they were predicted. We want to point out that they are not
single values, but rather continued dimensions that could be outlined
differently depending on the study system. Regardless of how they were
defined, they always delineate a specific area \(A\) and duration \(t\).
These could refer to the sampled area and duration or to the targeted
location and time period.

Many factors could be taken into consideration when estimating the
probability that a predator species \(i\) interacts with a given prey
species \(j\) locally. One of the most important is species
co-occurrence \(C\), which is a Boolean describing if both species can
be found at location and time (\(x\), \(y\), \(t\)). Surely, the
probability that the interaction is realized must be \(0\) when species
do not co-occur (\(C = 0\)). Interaction probabilities can also be
conditional on other biological and environmental variables, such as
temperature, precipitation, presence of shelters, phenotypic plasticity,
phenology, and presence of other interacting species in the network.
These conditions can affect species traits, which greatly impact the
capacity of species to interact (Poisot \emph{et al.} 2015). Similarly,
species relative abundance is another important predictor of the
probability of interaction, because it impacts the probability that
species will randomly encounter (Canard \emph{et al.} 2012; Canard
\emph{et al.} 2014; Poisot \emph{et al.} 2015). Here, we will use the
variable \(\Omega\) as a substitute for the biological and ecological
context in which interaction probabilities were estimated. This gives us
the following equation for the probability of realized interaction
between species (or taxa) \(i\) and \(j\) in a local network \(N\):

\begin{equation}\protect\hypertarget{eq:local}{}{P_{N}(i \rightarrow j | A, t, C, \Omega),}\label{eq:local}\end{equation}

which can be read as the probability of local interaction between the
two species in an area \(A\) and time interval \(t\), given their
co-occurrence \(C\) and specific environmental and biological conditions
\(\Omega\).

Multiple difficulties of interpretation arise when the conditions are
not clearly specified, which we found is often the case in the
literature. For example, if \(P_{N}(i \rightarrow j | C = 1)\)
represents the probability that two co-occurring species interact (i.e.,
the edge's probability value), \(P_{N}(i \rightarrow j)\) denotes
instead the probability of interaction without knowing if they co-occur
(i.e., the product of the nodes and edge's probability values). For
practical reasons, probabilistic ecological networks are generally
represented as matrices of probabilities (i.e., matrices of edges
without node values), whose elements are thus hard to interpret without
clear indications about \(C\). Overall, when probabilities of
interactions are estimated using specific values of \(A\), \(t\), \(C\),
and \(\Omega\), ecologists should make them explicit in their metadata,
preferably using mathematical equations to avoid any ambiguity. Below,
we will see examples of why this matters when it comes to spatial,
temporal, and taxonomic scaling of biotic interactions.

@@

Even though the spatiotemporal variability of interactions is not
considered in metawebs, they can still be useful to reconstruct local
networks of realized interactions. Indeed, local networks are formed
from subsets of their metaweb (called subnetworks), which are obtained
by selecting a subset of both species and interactions (Dunne 2006).
Because a community's composition is arguably easier to sample (or
predict) than its interactions, the biggest challenge is to sample links
from the metaweb. This becomes a conceptual issue when we consider how
potential and realized interactions differ. Despite these concerns,
metawebs remain an important source of ecological information that can
be leveraged for realistically predicting spatially explicit networks.
First, metawebs set the upper limit of species interactions (McLeod
\emph{et al.} 2021), i.e. the probability that two species interact at a
specific location is always lower or equal to the probability of their
potential interaction:

\begin{equation}\protect\hypertarget{eq:switch}{}{P_{N}(i \rightarrow j | A, t, C, \Omega) \le
P_M(i \rightarrow j).}\label{eq:switch}\end{equation}

Therefore, inferring local networks from their metaweb keeping the same
values of interaction probability would generate systematic biases in
the prediction. In that case, these networks would instead be called
\emph{spatially explicit} or \emph{local} metawebs (i.e., smaller-scale
networks of potential interactions). Second, the structure of local
networks is constrained by the one of their metaweb (Saravia \emph{et
al.} 2022). This suggests that a metaweb not only constrains the
pairwise interactions of its corresponding local networks, but also
their emerging properties. Inferring the structure of local networks
from the metaweb could thus help estimate more realistically the
likelihood that potential interactions are realized and observed locally
(Strydom \emph{et al.} 2021).

@@

\begin{itemize}
\tightlist
\item
  Why do probabilistic local food webs scale with area and time but not
  probabilistic metawebs?
\item
  In metawebs, interaction probabilities do not scale with space and
  time because they depend solely on the biological capacity of two
  species to interact.
\item
  In local food webs, interaction probabilities scale with space and
  time because there are more opportunities of interactions (e.g., more
  environmental conditions) in a larger area and longer time period.
\item
  What are some network area relationships in probabilistic local food
  webs?
\item
  We know that local networks can inform regional networks. However, can
  regional networks inform local networks?
\item
  Sampling local networks from metawebs vs from probabilistic local
  networks
\end{itemize}

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

The emergence of probabilistic thinking in network ecology has paved the
way to a better assessment of the spatiotemporal variability and
uncertainty of biotic interactions. However, we showed that network
metadata are perhaps as important as interaction data themselves when it
comes to interpreting probabilistic food webs in ecological terms.
First, the type of probabilistic interaction (potential or realized)
impacts the importance of scale, with interactions in metawebs being
scale independent, both spatially and temporally. Second, the conditions
in which local networks were estimated (e.g., area, time, biological and
environmental factors) and the attributes of the interacting species
that were considered (e.g., species co-occurrence) are essential
contextual factors that impact the mathematical representation of
probabilities and their resulting behaviour. Third, the biological
interpretation of probabilities changes with the level of organization
of the network due to the aggregation of individuals into broader
groups. All these information should be available as clear metadata so
that ecologists can use probabilistic network data appropriately.

Moreover, many statistical models in ecology that yield accurate
predictions of biotic interactions are black boxes difficult to
interpret. Ecologists should be careful before using the output of these
models as probabilistic objects, since there is often a thin line
between a real probability and a non-probabilistic predictive number (or
score). Probabilities are numbers between \(0\) and \(1\) that sum to
\(1\) and either represent the expected frequency of a phenomenon or the
degree of belief that it will be realized. Non-probabilistic scores,
which are more akin to interaction weights, have different mathematical
properties, which impacts how we should handle these numbers in a
spatially or temporally explicit context. Therefore, researchers should
use their expertise to assess if their interaction data are actually
probabilities or scores. This should also be added to the metadata
before sharing them, as well as the methods used to build the networks.

Better metadata documentation would allow researchers to use and
manipulate probabilistic ecological interactions according to how they
were actually defined and obtained. This would support better scientific
practices, in particular when these data are used for ecological
prediction and forecasting. For instance, we showed that building a
rigorous workflow to predict local networks from a probabilistic metaweb
requires a good understanding of the data at hand. Similarly, explicitly
stating the context in which probabilistic data were estimated would
help using forecasting food-web models more rigorously under specific
climate change and habitat use scenarios. Regardless of the method and
application, fostering a better foundation for probabilistic reasonings
in network ecology, from the very nature of probabilities and biotic
interactions, is essential.

@@

\leavevmode\hypertarget{ref-Pringle2020ResFoo}{}%
Pringle, R.M. \& Hutchinson, M.C. (2020). Resolving Food-Web Structure.
\emph{Annual Review of Ecology, Evolution, and Systematics}, 51, 55--80.
